On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-08-24 paper.tex
Inverse sequence in OEIS
Minor corrections ("t" replaced by "\tiles"); and clarifications

From Theorem~\ref{theo: funt, sq, form} it is easily seen that odd values of $\funt(\len)$ correspond to $\isoli = \jsoli$, whereas even values are achieved with $\jsoli = \isoli+1$. Given $\tiles \geq 3$ with $t$ odd, $\funt(\len) = \tiles$ if and only
\begin{equation}
\label{eq: square, odd max tiles, lengths}
\frac \len a \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{ \sqrt{(\tiles-1)^2+(\tiles-3)^2}}{2} \right].
\end{equation}
For $\tiles \geq 4$, $t$ even, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, even max tiles, lengths}
\frac \len a \in \left(\frac{ \sqrt{(\tiles-2)^2+(\tiles-4)^2}}{2}, \frac{\tiles-2}{\sqrt{2}} \right].
\end{equation}

@@

The sequence $\funli(\tiles)$, $\tiles \in \mathbb N$ can be computed using \eqref{eq: theo: funt, sq, form; funl} and \eqref{eq: funli funl}, and has initial values $1, 1, 1, 2, 2, 3, 3, 4, 5 \ldots$, as shown in Figure~\ref{fig: funli}. This is A**** in the On-Line Encyclopedia of Integer Sequences \cite{***}. However, a simpler expression can be obtained from \eqref{eq: funti leni} and \eqref{eq: funli funti}. This is established by the next theorem, which also states some properties of $\funli$, parallel to those of $\funti$.

@@

\section{Arbitrary grid with real-valued lengths}

@@

Each realization of the random segment gives rise to a canonical rectangle. The normalized dimensions $i$ and $j$ of the canonical rectangle are thus random variables. Except for a set of realizations with zero probability, $i$ and $j$ are at least $1$, and the segment visits $t = i+j-1$ tiles.

The random variables $i$ and $j$ are not independent. As a consequence, while the marginal distributions of $i$ and $j$ are easy to compute, the distribution of $i+j$ is difficult. However, knowing the marginal distributions is sufficient to compute the average number of visited tiles:
\begin{equation}
\label{eq: funta E[i] E[j]}
\funta(\len) = \E[t] = \E[i]+\E[j]-1.
\end{equation}

@@

In view of Theorem~\ref{theo: funta, form}, the average number of visited tiles has a remarkably simple dependence on the segment length, namely an affine function. Conversely, for any $t>1$ it is immediate to compute the length of a random segment that visits $t$ tiles on average, given as $\funta^{-1}(t)$.

@@

***Clarify here that computing the probability of visiting the maximum number of tiles in the general case is a difficult problem, and it will be dealt with in the square-grid case only

@@

\begin{theorem}
\label{theo: probmax, sq}
Given $a, \len \in \mathbb R^+$, consider a square grid with spacing $a$ and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles $\funt(\len)$ is
\begin{equation}
\label{eq: probmax}
\begin{split}
\probmax(\len) & = g\left(\frac{\len} a, 
\tiles-3, \tiles-3 \right) \quad\ \text{for } t \text{ odd, } \textstyle \frac \len a \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = g\left(\frac{\len} a, \tiles-3, \tiles-3 \right) + 2 g\left(\frac{\len} a, \tiles-5, \tiles-1 \right) \\
& \text{for } t \text{ odd, } \textstyle \frac \len a \in \left (\frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2,  \frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\frac{\len} a, \tiles-4, \tiles-2 \right) \quad \text{for } t \text{ even, } \textstyle \frac \len a \in \left( \frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} 2, \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\frac{\len} a, \tiles-4, \tiles-2 \right) + 2 g\left(\frac{\len} a, \tiles-6, \tiles \right) \\
& \text{for } t \text{ even, } \textstyle \frac \len a \in \left( \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2, \frac{\tiles-2}{\sqrt{2}} \right].
\end{split}
\end{equation}
\end{theorem}

@@

\begin{figure}
\centering%
\subfigure[The second endpoint is in the tile with lower-left corner $((t-1)/2,(t-1)/2)$]{%
\label{fig: probmax_funt_odd_diag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_diag}%
}\\%\hfill%
\subfigure[The second endpoint is in the tile with lower-left corner $((t-3)/2,(t+1)/2)$]{%
\label{fig: probmax_funt_odd_offdiag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_offdiag}%
}%
\caption{Segment orientations and positions of the first endpoint within the reference tile for a given number $\tiles$ of visited tiles. Example for odd $\tiles$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

@@

The situation is different depending on whether $\tiles$ is odd or even. Consider $\tiles$ odd first. Under this assumption, the range of lengths corresponding to $\tiles$ is given by \eqref{eq: square, odd max tiles, lengths}. For any length in this interval, there are some positions within the reference file from which the tile with lower-left coordinate $((t-1)/2, (t-1)/2)$ can be reached. Equivalently, the shifted segment with one endpoint at $((t-1)/2, (t-1)/2)$ can reach the reference tile. This is illustrated in Figure~\ref{fig: probmax_funt_odd_diag}. In addition, if 
\begin{equation}
\label{eq: probmax len odd2}
\len > \frac{\sqrt{(\tiles-5)^2 + (\tiles-1)^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $((t-3)/2, (t+1)/2)$ and $((t+1)/2, (t-3)/2)$ are also reachable from the reference tile; or equivalently the shifted segment can reach the reference tile, as shown in Figure~\ref{fig: probmax_funt_odd_offdiag}. This can only happen for $\tiles \geq 5$ (for $\tiles=3$ the inequality \eqref{eq: probmax len odd2} cannot be satisfied for lengths in the range given by \eqref{eq: square, odd max tiles, lengths}). In this case, by symmetry, each of those two tiles contributes the same probability. Other tiles corresponding to the same $\tiles$ need not be considered, because they cannot be reached with lengths satisfying \eqref{eq: square, odd max tiles, lengths}.

For $\tiles$ even, which corresponds to lengths in the interval \eqref{eq: square, even max tiles, lengths}, the two tiles with lower-left coordinates $(t/2-1, t/2)$ and $(t/2, t/2-1)$ can always be reached from some positions within the reference tile. In addition, if
\begin{equation}
\label{eq: probmax len even2}
\len > \frac{\sqrt{((\tiles-6)^2 + \tiles^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $(\tiles/2-2, \tiles/2+1)$ and $(\tiles/2+1, \tiles/2-2)$ have to be considered too. This can only happen for $\tiles \geq 8$. No other tiles can be reached from the reference tile.

@@

In the first case, as has been argued, only the tile with lower-left coordinates $((t-1)/2, (t-1)/2)$ is reachable. For a given $\theta$, the conditional probability $\probmax_{(t-1)/2, (t-1)/2}(\len, \theta)$ is the area of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_diag} divided by $a^2$. The horizontal and vertical sides of this rectangle are $\len\sin\theta - a(\tiles-3)/2$ and $\len\cos\theta - a(\tiles-3)/2$ respectively, and thus
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{(t-1)/2, (t-1)/2}(\len, \theta) = \frac{(\len\sin\theta - a(\tiles-3)/2) (\len\cos\theta - a(\tiles-3)/2)} {a^2}
\end{equation}
The range of allowed values for $\theta$, as can be seen from Figure~\ref{fig: probmax_funt_odd_diag}, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{a(\tiles-3)}{2\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{a(\tiles-3)}{2\len}.
\end{align}
Computing the average of \eqref{eq: probmax: probmax theta odd} over $\theta$, multiplying by $4$ to include all quadrants and comparing with \eqref{eq: g def} gives $\probmax(\len)$ as
\begin{equation}
\label{eq: probmax: probmax odd}
\begin{split}
\probmax(\len) & = 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(t-1)/2, (t-1)/2}(\len, \theta)}{2\pi} \diff \theta \\
& = \frac 2 \pi \int_{\arcsin\frac{a(\tiles-3)}{2\len}}^{\arccos\frac{a(\tiles-3)}{2\len}} \left( \frac{\len\sin\theta}a - \frac{\tiles-3} 2 \right)  \left( \frac{\len\cos\theta}a - \frac{\tiles-3} 2 \right) \diff \theta \\
& = g\left( \frac \len a, t-3, t-3 \right ),
\end{split}
\end{equation}
in accordance with \eqref{eq: probmax}.

For the second case in \eqref{eq: probmax}, $\probmax(\len)$ is obtained as in the preceding paragraph plus an additional term corresponding to the contribution of the tile with lower-left corner $((t-3)/2, (t+1)/2)$, multiplied by $2$ to account for its symmetrical tile $((t+1)/2, (t-3)/2)$ and by $4$ to include all quadrants. The sides of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_offdiag} are $\len\sin\theta - a(\tiles-5)/2$ and $\len\cos\theta - a(\tiles-1)/2$, and therefore
\begin{equation}
\label{eq: probmax: probmax theta odd2}
\probmax_{(t-3)/2, (t+1)/2}(\len, \theta) = \frac{(\len\sin\theta - a(\tiles-5)/2) (\len\cos\theta - a(\tiles-1)/2)} {a^2}.
\end{equation}
The integration interval $(\theta_0, \theta_1)$ is in this case
\begin{align}
\theta_0 & = \arcsin\frac{a(\tiles-5)}{2\len}, \\
\theta_1 & = \arccos\frac{a(\tiles-1)}{2\len}.
\end{align}
The additional contribution to $\probmax(\len)$, given by
\begin{multline}
8 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(t-3)/2, (t+1)/2}(\len, \theta)}{2\pi} \diff \theta \\
= \frac 4 \pi \int_{\arcsin\frac{a(\tiles-5)}{2\len}}^{\arccos\frac{a(\tiles-1)}{2\len}} \left( \frac{\len\sin\theta}a - \frac{\tiles-5} 2 \right)  \left( \frac{\len\cos\theta}a - \frac{\tiles-1} 2 \right) \diff \theta,
\end{multline}
is recognized to be $2 g(\len/a, \tiles-5, \tiles-1)$. This proves the second equality in \eqref{eq: probmax}.

@@

>>>>

From Theorem~\ref{theo: funt, sq, form} it is easily seen that odd values of $\funt(\len)$ correspond to $\isoli = \jsoli$, whereas even values are achieved with $\jsoli = \isoli+1$. Given $\tiles \geq 3$ with $\tiles$ odd, $\funt(\len) = \tiles$ if and only
\begin{equation}
\label{eq: square, odd max tiles, lengths}
\frac \len a \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{ \sqrt{(\tiles-1)^2+(\tiles-3)^2}}{2} \right].
\end{equation}
For $\tiles \geq 4$, $\tiles$ even, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, even max tiles, lengths}
\frac \len a \in \left(\frac{ \sqrt{(\tiles-2)^2+(\tiles-4)^2}}{2}, \frac{\tiles-2}{\sqrt{2}} \right].
\end{equation}

@@

The sequence $\funli(\tiles)$, $\tiles \in \mathbb N$ can be computed using \eqref{eq: theo: funt, sq, form; funl} and \eqref{eq: funli funl}, and has initial values $1, 1, 1, 2, 2, 3, 3, 4, 5 \ldots$, as shown in Figure~\ref{fig: funli}. This is A346693 in the On-Line Encyclopedia of Integer Sequences \cite{OEIS_unitsq_int_inv}. However, a simpler expression can be obtained from \eqref{eq: funti leni} and \eqref{eq: funli funti}. This is established by the next theorem, which also states some properties of $\funli$, parallel to those of $\funti$.

@@

Each realization of the random segment gives rise to a canonical rectangle, whose normalized dimensions $i$ and $j$ are thus random variables. Except for a set of realizations with zero probability, $i$ and $j$ are at least $1$, and the segment visits $i+j-1$ tiles. Note that $i$ and $j$ are not statistically independent.

\subsection{Arbitrary grid with real-valued lengths}

@@

Although the random variables $i$ and $j$ are dependent, their marginal distributions can be obtained easily, and are sufficient to compute the average number of visited tiles:
\begin{equation}
\label{eq: funta E[i] E[j]}
\funta(\len) = \E[\tiles] = \E[i]+\E[j]-1.
\end{equation}

@@

In view of Theorem~\ref{theo: funta, form}, the average number of visited tiles has a remarkably simple dependence on the segment length, namely an affine function. Conversely, for any $\tiles>1$ it is immediate to compute the length of a random segment that visits $\tiles$ tiles on average, given as $\funta^{-1}(\tiles)$.

@@

The probability of visiting the maximum number of tiles for a rectangular grid is difficult to compute, due to the irregularity of the minimal sufficient set (discussed in \S\ref{part: fund results}). In the square case, however, the simplicity of this set makes the problem tractable.

@@

\begin{theorem}
\label{theo: probmax, sq}
Given $a, \len \in \mathbb R^+$, consider a square grid with spacing $a$ and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles $\funt(\len)$ is
\begin{equation}
\label{eq: probmax}
\begin{split}
\probmax(\len) & = g\left(\frac{\len} a, 
\tiles-3, \tiles-3 \right) \quad\ \text{for } \tiles \text{ odd, } \textstyle \frac \len a \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = g\left(\frac{\len} a, \tiles-3, \tiles-3 \right) + 2 g\left(\frac{\len} a, \tiles-5, \tiles-1 \right) \\
& \text{for } \tiles \text{ odd, } \textstyle \frac \len a \in \left (\frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2,  \frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\frac{\len} a, \tiles-4, \tiles-2 \right) \quad \text{for } \tiles \text{ even, } \textstyle \frac \len a \in \left( \frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} 2, \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\frac{\len} a, \tiles-4, \tiles-2 \right) + 2 g\left(\frac{\len} a, \tiles-6, \tiles \right) \\
& \text{for } \tiles \text{ even, } \textstyle \frac \len a \in \left( \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2, \frac{\tiles-2}{\sqrt{2}} \right].
\end{split}
\end{equation}
\end{theorem}

@@

\begin{figure}
\centering%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-1)/2,(\tiles-1)/2)$]{%
\label{fig: probmax_funt_odd_diag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_diag}%
}\\%\hfill%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-3)/2,(\tiles+1)/2)$]{%
\label{fig: probmax_funt_odd_offdiag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_offdiag}%
}%
\caption{Segment orientations and positions of the first endpoint within the reference tile for a given number $\tiles$ of visited tiles. Example for odd $\tiles$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

@@

The situation is different depending on whether $\tiles$ is odd or even. Consider $\tiles$ odd first. Under this assumption, the range of lengths corresponding to $\tiles$ is given by \eqref{eq: square, odd max tiles, lengths}. For any length in this interval, there are some positions within the reference file from which the tile with lower-left coordinate $((\tiles-1)/2, (\tiles-1)/2)$ can be reached. Equivalently, the shifted segment with one endpoint at $((\tiles-1)/2, (\tiles-1)/2)$ can reach the reference tile. This is illustrated in Figure~\ref{fig: probmax_funt_odd_diag}. In addition, if 
\begin{equation}
\label{eq: probmax len odd2}
\len > \frac{\sqrt{(\tiles-5)^2 + (\tiles-1)^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $((\tiles-3)/2, (\tiles+1)/2)$ and $((\tiles+1)/2, (\tiles-3)/2)$ are also reachable from the reference tile; or equivalently the shifted segment can reach the reference tile, as shown in Figure~\ref{fig: probmax_funt_odd_offdiag}. This can only happen for $\tiles \geq 5$ (for $\tiles=3$ the inequality \eqref{eq: probmax len odd2} cannot be satisfied for lengths in the range given by \eqref{eq: square, odd max tiles, lengths}). In this case, by symmetry, each of those two tiles contributes the same probability. Other tiles corresponding to the same $\tiles$ need not be considered, because they cannot be reached with lengths satisfying \eqref{eq: square, odd max tiles, lengths}.

For $\tiles$ even, which corresponds to lengths in the interval \eqref{eq: square, even max tiles, lengths}, the two tiles with lower-left coordinates $(\tiles/2-1, \tiles/2)$ and $(\tiles/2, \tiles/2-1)$ can always be reached from some positions within the reference tile. In addition, if
\begin{equation}
\label{eq: probmax len even2}
\len > \frac{\sqrt{((\tiles-6)^2 + \tiles^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $(\tiles/2-2, \tiles/2+1)$ and $(\tiles/2+1, \tiles/2-2)$ have to be considered too. This can only happen for $\tiles \geq 8$. No other tiles can be reached from the reference tile.

@@

In the first case, as has been argued, only the tile with lower-left coordinates $((\tiles-1)/2, (\tiles-1)/2)$ is reachable. For a given $\theta$, the conditional probability $\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)$ is the area of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_diag} divided by $a^2$. The horizontal and vertical sides of this rectangle are $\len\sin\theta - a(\tiles-3)/2$ and $\len\cos\theta - a(\tiles-3)/2$ respectively, and thus
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta) = \frac{(\len\sin\theta - a(\tiles-3)/2) (\len\cos\theta - a(\tiles-3)/2)} {a^2}
\end{equation}
The range of allowed values for $\theta$, as can be seen from Figure~\ref{fig: probmax_funt_odd_diag}, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{a(\tiles-3)}{2\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{a(\tiles-3)}{2\len}.
\end{align}
Computing the average of \eqref{eq: probmax: probmax theta odd} over $\theta$, multiplying by $4$ to include all quadrants and comparing with \eqref{eq: g def} gives $\probmax(\len)$ as
\begin{equation}
\label{eq: probmax: probmax odd}
\begin{split}
\probmax(\len) & = 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)}{2\pi} \diff \theta \\
& = \frac 2 \pi \int_{\arcsin\frac{a(\tiles-3)}{2\len}}^{\arccos\frac{a(\tiles-3)}{2\len}} \left( \frac{\len\sin\theta}a - \frac{\tiles-3} 2 \right)  \left( \frac{\len\cos\theta}a - \frac{\tiles-3} 2 \right) \diff \theta \\
& = g\left( \frac \len a, \tiles-3, \tiles-3 \right ),
\end{split}
\end{equation}
in accordance with \eqref{eq: probmax}.

For the second case in \eqref{eq: probmax}, $\probmax(\len)$ is obtained as in the preceding paragraph plus an additional term corresponding to the contribution of the tile with lower-left corner $((\tiles-3)/2, (\tiles+1)/2)$, multiplied by $2$ to account for its symmetrical tile $((\tiles+1)/2, (\tiles-3)/2)$ and by $4$ to include all quadrants. The sides of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_offdiag} are $\len\sin\theta - a(\tiles-5)/2$ and $\len\cos\theta - a(\tiles-1)/2$, and therefore
\begin{equation}
\label{eq: probmax: probmax theta odd2}
\probmax_{(\tiles-3)/2, (\tiles+1)/2}(\len, \theta) = \frac{(\len\sin\theta - a(\tiles-5)/2) (\len\cos\theta - a(\tiles-1)/2)} {a^2}.
\end{equation}
The integration interval $(\theta_0, \theta_1)$ is in this case
\begin{align}
\theta_0 & = \arcsin\frac{a(\tiles-5)}{2\len}, \\
\theta_1 & = \arccos\frac{a(\tiles-1)}{2\len}.
\end{align}
The additional contribution to $\probmax(\len)$, given by
\begin{multline}
8 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(\tiles-3)/2, (\tiles+1)/2}(\len, \theta)}{2\pi} \diff \theta \\
= \frac 4 \pi \int_{\arcsin\frac{a(\tiles-5)}{2\len}}^{\arccos\frac{a(\tiles-1)}{2\len}} \left( \frac{\len\sin\theta}a - \frac{\tiles-5} 2 \right)  \left( \frac{\len\cos\theta}a - \frac{\tiles-1} 2 \right) \diff \theta,
\end{multline}
is recognized to be $2 g(\len/a, \tiles-5, \tiles-1)$. This proves the second equality in \eqref{eq: probmax}.

@@

It should be noted that the technique used in the proof of Theorem~\ref{theo: probmax, sq} could in principle be used for obtaining the probability that the number of tiles visited by the segment equals or exceeds any given value $t \funt(\len)$. The resulting expressions, however, are cumbersome, because more cases (more subintervals for the length) need to be distinguished depending on how many tiles with $i+j-1=t$ can be reached from the reference tile.
